The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2025-04-22 Paper.qmd
Edits after full read through.

Although infant signals of need, particularly infant cries, have been extensively studied in many populations [@barr1991; @fouts2005; @konner1972; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, how and when older children signal need has rarely been investigated [for one exception, see @syme2023]. The aim of this study was to help fill these gaps by investigating three common types of child signals.

@@

Although the fitness prospects of children and their familial caretakers are largely intertwined, evolutionary theory predicts that disagreements about the amount of investment from parents and other close kin will be ubiquitous [@fouts2005; @godfray1995; @godfray1992; @trivers1974; @bossan2013] because each child shares more genes through descent with themselves (100%) than with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974; @bossan2013].

For the same reason, siblings also often find themselves in conflicts of interest with one another, with each expected to prefer more resources for himself or herself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, and large numbers of siblings have been associated with decreased growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

@@

Humans are a uniquely interdependent species among the living primates [@tomasello2012] in that both sexes' fitness prospects depend greatly on a network of close kin and other social allies [@apicella2019]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from mothers and fathers [@kramer2018]. However, grandparents, aunts, uncles, and non-relatives also provide significant investment in children in many societies [@kramer2018; @sear2008], which was likely favored due to a combination of inclusive fitness benefits, reciprocity, and reputation management [@gurven2000; @schniter2023]. Children also often serve as alloparents, helping to raise their younger siblings via direct care, food collection and preparation, and other family tasks [@hrdy2009; @meehan2009; @sear2008] despite still requiring substantial amounts of investment themselves [@kramer2018a].

@@

When the fitness interests of signalers and receivers are aligned, i.e., there are no conflicts of interest, child signals of need are expected to be low cost, what @maynardsmith1995 refer to as "minimal signals," those whose cost is no greater than that needed to transmit the information effectively. In other words, although minimal signals might involve some cost (e.g., energy for a loud scream), they do not require any additional mechanism to enforce honesty.

@@

In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. When there are conflicts of interest between children and their caregivers, children might need to use signals whose costs outweigh the benefits of additional care for those not in need, but would result in a net benefit to those with genuine needs. In doing so, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing. When this is the case, all parties are better off than they would be without the signal, with caretakers benefitting from the revelation of fitness-relevant information and the child benefitting from the increased likelihood of a beneficial response.

However, even credible signals of need might not result in a beneficial response because belief in a child’s need is not the only factor determining the form and amount of support given by caregivers. Instead, caregivers are expected to weigh the costs and benefits of support against other options [@daly1987; @daly1983]. Ancestrally, this likely often meant evaluating the viability of the child, which favored selection for child signals of vigor [@soltis2004]. In general, selection favors allocation of parental investment to current and future offspring that equalizes marginal fitness returns to the parent [@clutton-brock1991; @daly1983; @dawkins1976; @hagen1999; @beaulieu2008; @frankenhuis2022; @daly1983].

Additionally, a parent may also decide against investing in a child, despite deeming that investment worthwhile, if time is not right (e.g., if a mother preferred to breastfeed during rest rather than when foraging). When this is the case, outward conflicts may exaggerate the extent of a conflict of interest and involve parents teaching children ways to behave that benefit them both [@altmann1980]. Finally, caregivers often have better information about the costs and benefits of various items and actions, and therefore can make more beneficial decisions for a child than the child can herself (e.g., preventing a child from playing with matches).

@@

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to signal need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. We conceptualize infrequent and transient sadness as a minimal signal of need. Frequent or prolonged sadness, however, is likely a form of withdrawal or depression, both of which might be effective costly signals of need when there are conflicts between the child and their caregivers [@gaffney2022; @syme2023].

@@

Although it is widely accepted that crying is a signal, there is no consensus on if or when it should be conceptualized as a minimal signal, a costly signal, an indexical signal, or a deceptive (manipulative) signal. Contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021], which would be type of minimal signal. Demonstrating vigor and health early in life was likely vital for much of our evolutionary history due to the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Children with severe medical conditions also produce acoustically distinct cries, suggesting that some elements of crying are indexical [@soltis2004]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models; @zahavi1977a; @lummaa1998; @eshel1991; @bergstrom1999 @kennedy2021].

@@

More specifically, we hypothesized that children have evolved strategies to tailor their signals of need to 1) their valuation of a given resource, 2) the presence of existing information about a child's needs and wants (i.e., potential to benefit from a resource) and 3) the extent of conflicts of interest with caretakers, which should influence the decision to use minimal signals or signals whose costs would deter deception. Such strategies can lead to some combination of verbal requests for more support and facial sadness, which we consider to be minimal signals, or withdrawal, temper tantrums and self-harm, which we consider "costly" signals. Crying might be considered a "minimal" signal in some circumstances, such as when a child needs a loud auditory signal to attract attention on a noisy playground, but a "costly" signal in others, such as when signaling vigor. Similarly, infrequent sadness would be a minimal signal, but frequent sadness might be better conceptualized as costly withdrawal or depression. Each of these might be effective in certain circumstances and are sometimes successful in eliciting more support from caretakers.

@@

This study of children aged `{r} paste(range(utila_df$ChildAge), collapse="-")` and their parents and other primary caretakers, was conducted in Utila, Honduras, one of the country’s small Bay Islands in the Western Caribbean. Participants in the study were sampled from two populations who face different levels of economic and social adversity: 1) a long-standing English-speaking community primarily comprising descendants of British immigrants who arrived on the island in the 1800’s, and 2) a more recent community of Spanish-speakers, mostly from mainland Honduras. Compared to the former, the latter are more likely to live in crowded and suboptimal conditions, face active discrimination and increased stress [@garcia2017], and experience economic uncertainty due the predominance of low paying, seasonal jobs primarily related to the island’s diving tourism industry [@garcia2018; @stonich2000]. Such inequality appears to be increasing in response to recent growth in the tourism industry, with the majority of wealth from tourism flowing to a small portion of native English-speaking islanders and other Westerners [@davis2014].

@@

To inform development of our survey questionnaire on child signaling, focus groups of Utilian parents were held in July 2022. Parents indicated frequent sibling competition and various ways children might signal need. Both Spanish-speaking and English-speaking moms reported that fighting between siblings was extremely common, with fights over toys, gifts, and phone time being the most reported causes. In both populations, crying and various forms of withdrawal were listed as the most common responses to a child not receiving investment or support, followed by temper tantrums and making messes in their home. Spanish-speaking mothers reported that refusing to do homework was a common tactic, while English-speaking mothers emphasized running away and self-harm (e.g., a child slamming their head on the ground) to a greater degree. In response to these hypothesized signals, discipline was common. Although the mothers generally reported that children were treated equally, needier children were described as getting more resources from parents, perhaps due to differences in signaling among children as well as information caretakers gained from cues of need.

@@

We used six-point Likert scales to get caregiver reports on the frequencies of 1) various types of child signaling, such as sadness and crying, 2) conflict between the caretaker and each participating child, 3) and child allocare of other children in their household. See @tbl-freqvars for exact wording. Caretakers were often not asked about child alloparenting behavior if a child was an only child. For this reason, we gave children an alloparenting frequency of “never” if they were the only child within a household. We also calculated the summed frequency of child alloparenting contributions by other children within the household.

@@

: Variable names and question wording for measures about the frequency of signaling, allocare, and conflict. All questions had the following labels attached to each point on its Likert scale: Never; Once a month or less; More than once a month but less than once a week; More than once a week but not daily; Daily; Multiple times per day. For analyses, we converted these categorical variables to numeric frequencies per month as follows: never = 0; once a month or less = 1; more than once a month = 3; more than once a week = 8; daily = 30; multiple times per day = 60. {#tbl-freqvars}

@@

Caretakers were also asked three questions using a five point Likert scale about their child's susceptibility to illness. We reversed the scores on two measures so that a score of 5 corresponded with the lowest susceptibility to illness before we averaged the scores of these to create a mean *illness avoidance* variable. In the `{r} sum((!is.na(utila_df$IllnessSusceptibility2)) & is.na(utila_df$IllnessSusceptibility3))` instance in which we only had data for 2 of the measures, we still calculated the mean but used a numerator of 2. See @stbl-healthvars for the wording of these variables.

@@

A)  Loadings of PC1 and PC2 on each variable. B) Biplot map of PC1 and PC2. Each point is one child. Arrows represent the loadings of PC1 and PC2 on each variable.
:::

@@

Minimum spanning tree derived from a Bayesian graphical model of child signaling and variables thought to cause differences in signaling strategies. Purple edges represent negative partial correlations between variables. Orange edges represent positive partial correlations between variables. Line weights represent partial correlation coefficients. Blue text represents our interpretation of the main clusters.
:::

@@

Child sadness, crying, and temper tantrums were common and highly correlated with one another. See @fig-barplot-SignalFreq and @fig-signal-corrplot. Caregivers reported that `{r} fmt_sum(utila_df$SadFreqN >= 8)` were sad at least once a week, `{r} fmt_sum(utila_df$CryFreqN >= 8)` cried at least once a week, and `{r} fmt_sum(utila_df$TantrumFreqN >= 8)` threw temper tantrums at least once a week. A small subset of children engaged in these signals/states at least daily; sadness: `{r} fmt_sum(utila_df$SadFreqN >= 30)`; crying: `{r} fmt_sum(utila_df$CryFreqN >= 30)`; and tantrums: `{r} fmt_sum(utila_df$TantrumFreqN >= 30)`. However, many children were reported to never engage in these signals/states; sadness: `{r} fmt_sum(utila_df$SadFreqN == 0)`; crying; `{r} fmt_sum(utila_df$CryFreqN == 0)`; and tantrums: `{r} fmt_sum(utila_df$TantrumFreqN == 0)`. For a visualization of how signal frequencies varied within children see @sfig-signal-alluvial-plot.

@@

Frequencies of child sadness, crying, and temper tantrums per month as reported by the primary caretaker. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

@@

Correlation matrix of the frequencies of signals/states covered in the study with sex, child age, and the frequency of conflict between children and primary caretakers. For a more complete correlation matrix of study variables see @sfig-plot-full-corrmat.
:::

@@

Lasso regression models of each child signaling frequency and conflict frequency. Variables whose coefficient values were shrunk to 0 are not displayed. Models fit using quasi-Poisson regression.
:::

@@

Effects of selected predictors of child signaling frequencies when holding continuous variables at their mean and factor variables at their mode.
:::

@@

Effects of selected predictors of child-parent conflict frequencies when holding continuous variables at their mean and factor variables at their mode.
:::

@@

The proportion of children perceived by caretakers as being more needy, as needy, or less needy than a caretaker's other children when holding continuous variables at their mean and factor variables at their mode. A: Effect of child age on relative neediness B: Effect child sadness frequency on relative neediness. Outcome variable question wording: How much time, effort, and resources does this child need compared to your other children?
:::

@@

Frequencies of caregiver response to the last instance of child signaling by the cause of the signaling. Categories of causes were based on cluster analysis of more fine-grained cause variables. See @sfig-cause-clustering for cluster dendrogram and @stbl-sumtab-causes for the frequencies of causes and caregiver responses.
:::

@@

Effects of selected predictors of the likelihood of positive, neutral, and negative responses to the last time a child became sad or cried when holding continuous variables at their mean and factor variables at their mode. 1 = evidence of the cause in the caregiver report; 0 = no evidence. The endpoints of the lines are dodged slightly for greater visibility. Dot size represents model predictions for each participant given their values of predictor variables. See @sfig-plot-caregiver-punish for the effect of the cause involving punishment.
:::

@@

Our bivariate correlations, PCA, MST, and lasso regression analyses all provided a consistent pattern of results: child age was a strong negative predictor of signaling frequency and cost; conflict with the primary caretaker was a positive predictor of signaling frequency and cost; and multiple indices of adversity, such as low neighborhood quality, low household income, and medical problems, were positive predictors of signaling frequency and cost (@fig-mst, @fig-PCA, @fig-signal-corrplot, @fig-lasso, @fig-signal-effects).

@@

["perceptions of need" instead of "perceived higher"?]

Consistent with the need signaling hypothesis, more frequent sadness was positively associated with perceived higher need relative to other children (@fig-effect-relativeneed). Higher perceived need, in turn, was associated with higher relative investment (@fig-plot-relative-investment). Crying and tantrums were not associated with higher perceived need, likely due to their high correlation with sadness and the propensity of lasso regression to retain only one of a set of correlated predictors [@zou2005]. Signaling behavior did not predict differences in relative caregiver investment beyond its effect on perceived need [@sfig-plot-invest-coefs], similar to an experimental vignette study that found that the likelihood of costly support was largely mediated by the effect of costly signaling on belief in need [@gaffney2022].

@@

Our results parallel those of previous studies of infants and young children that found that, outside of peaks in infancy and early childhood, the frequencies of crying and temper tantrums decrease with age [@macfarlane1954; @ward1970; @osterman2010; @vermillet2022; @barr1991; @potegal2003; @bhatia1990]. Studies examining crying and tantrums in later childhood and adolensce are rare, especially for crying [@rottenberg2012; @zeifman2013]. Our findings suggest the frequency of crying continues to decrease into childhood, dropping precipitously as children grow older. The frequency of temper tantrums also declined, as did the frequency of child sadness, albeit at a much slower rate (@fig-lasso, @fig-signal-effects). Our results also align with studies examining responses to the crying and fussing of infants that have shown that they often lead to increased support [e.g., breastfeeding, attention, or maintaining proximity; @lin2012; @kushnick2006; @kruger2010; @hewlett1998; @chaudhary2024; @tronick1987], but can also lead  to negative responses like abuse [@zeevenhooven2018; @kurth2011; @howard2006; @soltis2004].

@@

One possible reason for the negative association of signaling with age, which is also seen in more industrialized settings [@zeifman2013; @potegal2003; @koch2003; @macfarlane1954], is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need [@osterman2010], and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while the energetic cost of crying might be costly for the youngest children, it is likely insignificant for older children. Older children might instead need to weigh the potential social costs of crying or otherwise expressing sadness [@glasberg1981; @glasberg1982; @zeman1996] against the potential increase in social support, which is consistent with our result that the age-related decline in signaling was steepest for crying. The age-related decline in signaling was also generally steeper for males than females [@sfig-signaltype].

@@

Although this study was broader than most in that it examined sadness, crying, and temper tantrums across much of childhood and adolescence, it did not cover all the ways that caregivers might infer need, such as cues of injury. It also did not investigate all potential signals of need. Pained facial and bodily expressions [@vigil2014; @craig2009; @cano2010; @schiavenato2010], self-harm outside of temper tantrums [@hagen2008; @nock2008], and suicidality [@syme2016] have all been suggested to credibly signal need in times of conflicts of interest and may be viable strategies for children, at least at certain ages. Running away has received similar treatment, especially in instances when it comes with substantial risk [@shostak1981] or is seen as a suicide-threat [@fischer1950; @syme2023]. For this reason, we also collected data on running away. However, mothers on Utila reported that running away on the island often involved children visiting neighbors or aunts and was much less risky than it would be on mainland Honduras, with a few mothers mentioning in focus groups that they warned their children about the difference in danger that came with running away on the mainland. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean and the presence of dogs, motorcycles, golf carts, and tuk-tuks (auto rickshaws) on the narrow roads typical of much of populated parts of the island. Due to uncertainty on how to conceptualize the risks involved, we did not include running away in our analyses.

>>>>

Although infant signals of need, particularly infant cries, have been extensively studied in many populations [@barr1991; @fouts2005; @konner1972; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, how and when older children signal need has rarely been investigated [for one exception, see @syme2023]. The aim of this study was to help fill these gaps by investigating three common types of signals of need in children and adolescents.

@@

Although the fitness prospects of children and their familial caretakers are largely intertwined, evolutionary theory predicts that disagreements about the amount of investment from parents and other close kin will be ubiquitous [@fouts2005; @godfray1995; @godfray1992; @trivers1974; @bossan2013] because each child shares more genes with themselves (100%) than (through recent common descent) with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974; @bossan2013].

For the same reason, siblings have both confluences and conflicts of interest with one another, with each expected to prefer more resources for himself or herself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, and large numbers of siblings have been associated with decreased growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

@@

Humans are a uniquely interdependent species among the living primates [@tomasello2012] in that both sexes' fitness prospects depend greatly on a network of close kin and other social allies [@apicella2019; @syme2024]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from mothers and fathers [@kramer2018]. However, grandparents, aunts, uncles, and non-relatives also provide significant investment in children in many societies [@kramer2018; @sear2008], which was likely favored due to a combination of inclusive fitness benefits, reciprocity, and reputation management [@gurven2000; @schniter2023]. Children also often serve as alloparents, helping to raise their younger siblings via direct care, food collection and preparation, and other family tasks [@hrdy2009; @meehan2009; @sear2008] despite still requiring substantial amounts of investment themselves [@kramer2018a].

@@

When the fitness interests of signalers and receivers are aligned, i.e., there are no conflicts of interest, child signals of need are expected to be what @maynardsmith1995 refer to as "minimal signals," signals whose cost is no greater than that needed to transmit the information effectively. In other words, although minimal signals might involve some cost (e.g., energy for a loud scream), they do not require any additional mechanism to enforce honesty.

@@

In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. When there are conflicts of interest between children and their caregivers, children might need to use signals whose costs outweigh the benefits of additional care for those not in need, but would result in a net benefit to those with genuine needs. In doing so, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing.

However, even credible signals of need might not result in a beneficial response because belief in a child’s need is not the only factor determining the form and amount of support given by caregivers. Instead, caregivers are expected to weigh the costs and benefits of support, which would take into account the reproductive value of current and future offspring, against other options [@daly1987; @daly1983], with selection favoring the allocation of parental investment to current and future offspring that equalizes marginal fitness returns to the parent [@clutton-brock1991; @daly1983; @dawkins1976; @hagen1999; @beaulieu2008; @frankenhuis2022; @daly1983].

Additionally, a parent may also decide against investing in a child, despite deeming that investment worthwhile, if the time is not right (e.g., if a mother preferred to breastfeed during rest rather than when foraging). When this is the case, outward conflicts might not reflect genuine conflicts of interest and instead involve parents teaching children ways to behave that benefit them both [@altmann1980]. Finally, caregivers often have better information about the costs and benefits of various items and actions, and therefore can make more beneficial decisions for a child than the child can herself (e.g., preventing a child from playing with matches).

@@

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to signal need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. We conceptualize infrequent and transient sadness signals as minimal signals of need [per @maynardsmith1995]. Frequent or prolonged sadness, however, is likely a form of withdrawal or depression, both of which might be effective costly signals of need when there are conflicts between the child and their caregivers [@gaffney2022; @syme2023].

@@

Although it is widely accepted that crying is a signal, there is no consensus on if or when it should be conceptualized as a minimal signal, a costly signal, an indexical signal, or a deceptive (manipulative) signal. Contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021], which would be a type of minimal signal. Demonstrating vigor and health early in life was likely vital for much of our evolutionary history due to the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Children with severe medical conditions also produce acoustically distinct cries, suggesting that some elements of crying are indexical [@soltis2004]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models, @zahavi1977a; @lummaa1998; @eshel1991; @bergstrom1999; @kennedy2021].

@@

More specifically, we hypothesized that children have evolved strategies to tailor their signals of need to 1) their valuation of a given resource, 2) the presence of existing information about a child's needs and wants (i.e., potential to benefit from a resource) and 3) the extent of conflicts of interest with caretakers, which should influence the decision to use minimal signals or signals whose costs would deter deception. Such strategies can lead to some combination of verbal requests for more support and facial sadness, which we consider to be minimal signals, or withdrawal, temper tantrums and self-harm, which we consider "costly" signals. Crying might be considered a "minimal" signal in some circumstances, such as when a child needs a loud auditory signal to attract attention on a noisy playground, but a "costly" signal in others, such as lengthy bouts that incur physiological costs in infants signaling vigor and social costs in older children signaling need. Similarly, infrequent sadness would be a minimal signal, but frequent sadness might be better conceptualized as costly withdrawal or depression.

@@

This study of children aged `{r} paste(range(utila_df$ChildAge), collapse="-")` and their parents and other primary caretakers was conducted in Utila, Honduras, one of the country’s small Bay Islands in the Western Caribbean. Participants in the study were sampled from two populations that face different levels of economic and social adversity: 1) a long-standing English-speaking community primarily comprising descendants of British immigrants who arrived on the island in the 1800’s, and 2) a more recent community of Spanish-speakers, mostly from mainland Honduras. Compared to the former, the latter are more likely to live in crowded and suboptimal conditions, face active discrimination and increased stress [@garcia2017], and experience economic uncertainty due the predominance of low paying, seasonal jobs primarily related to the island’s diving tourism industry [@garcia2018; @stonich2000]. Such inequality appears to be increasing in response to recent growth in the tourism industry, with the majority of wealth from tourism flowing to a small portion of native English-speaking islanders and other Westerners [@davis2014].

@@

To inform development of our survey questionnaire on child signaling, focus groups of Utilian parents were held in July 2022. Parents indicated frequent sibling competition and various ways children might signal need. Both Spanish-speaking and English-speaking moms reported that fighting between siblings was extremely common, with fights over toys, gifts, and phone time being the most reported causes. In both populations, crying and various forms of sadness and withdrawal were listed as the most common responses to a child not receiving investment or support, followed by temper tantrums. In response to these hypothesized signals, discipline was common. Although the mothers generally reported that children were treated equally, needier children were described as getting more resources from parents, perhaps due to differences in signaling among children as well as information caretakers gained from cues of need.

@@

We used six-point Likert scales to get caregiver reports on the frequencies of 1) various types of child signaling, such as sadness and crying, 2) conflict between the caretaker and each participating child, 3) and child allocare of other children in their household. See @tbl-freqvars for exact wording. We gave children an alloparenting frequency of “never” if they were the only child within a household. We also calculated the summed frequency of child alloparenting contributions by other children within the household.

@@

: Variable names and question wording for measures about the frequency of signaling, allocare, and conflict. All questions had the following labels attached to each point on its Likert scale: Never; Once a month or less; More than once a month but less than once a week; More than once a week but not daily; Daily; Multiple times per day. For analyses, we converted these categorical variables to numeric frequencies per month as follows: never = 0; once a month or less = 1; more than once a month = 3; more than once a week = 8; daily = 30; multiple times per day = 60 {#tbl-freqvars}

@@

Caretakers were also asked three questions using a five point Likert scale about their child's susceptibility to illness. We reversed the scores on two measures so that a score of 5 corresponded with the lowest susceptibility to illness before we averaged the scores of these to create a mean *illness avoidance* variable. In the `{r} sum((!is.na(utila_df$IllnessSusceptibility2)) & is.na(utila_df$IllnessSusceptibility3))` instance in which we only had data for two of the measures, we still calculated the mean but used a numerator of 2. See @stbl-healthvars for the wording of these variables.

@@

a) Loadings of each variable on PC1 and PC2. b) Biplot of PC1 and PC2. Each point is one child. Arrows represent the loadings of each variable on PC1 and PC2. Variables with small loading factors not displayed
:::

@@

Minimum spanning tree derived from a Bayesian graphical model of child signaling and variables thought to cause differences in signaling strategies. Purple edges represent negative partial correlations between variables. Orange edges represent positive partial correlations between variables. Line weights represent partial correlation coefficients. Blue text represents our interpretation of the main clusters
:::

@@

Child sadness, crying, and temper tantrums were common and highly correlated with one another. See @fig-barplot-SignalFreq and @fig-signal-corrplot. Caregivers reported that `{r} fmt_sum(utila_df$SadFreqN >= 8)` were sad at least once a week, `{r} fmt_sum(utila_df$CryFreqN >= 8)` cried at least once a week, and `{r} fmt_sum(utila_df$TantrumFreqN >= 8)` threw temper tantrums at least once a week. A small subset of children engaged in these signals/states at least daily; sadness: `{r} fmt_sum(utila_df$SadFreqN >= 30)`; crying: `{r} fmt_sum(utila_df$CryFreqN >= 30)`; and tantrums: `{r} fmt_sum(utila_df$TantrumFreqN >= 30)`. However, many children were reported to never engage in these signals/states; sadness: `{r} fmt_sum(utila_df$SadFreqN == 0)`; crying; `{r} fmt_sum(utila_df$CryFreqN == 0)`; and tantrums: `{r} fmt_sum(utila_df$TantrumFreqN == 0)`. For a visualization of how signal frequencies varied within children, see @sfig-signal-alluvial-plot.

@@

Frequencies of child sadness, crying, and temper tantrums per month as reported by the primary caretaker. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day"
:::

@@

Correlation matrix of the frequencies of signals/states covered in the study with sex, child age, and the frequency of conflict between children and primary caretakers. For a more complete correlation matrix of study variables see @sfig-plot-full-corrmat
:::

@@

Lasso regression models of each child signaling frequency and conflict frequency. Variables whose coefficient values were shrunk to 0 are not displayed. Models fit using quasi-Poisson regression
:::

@@

Effects of selected predictors of child signaling frequencies when holding continuous variables at their mean and factor variables at their mode
:::

@@

Effects of selected predictors of child-parent conflict frequencies when holding continuous variables at their mean and factor variables at their mode
:::

@@

The proportion of children perceived by caretakers as being more needy, as needy, or less needy than a caretaker's other children when holding continuous variables at their mean and factor variables at their mode. A: Effect of child age on relative neediness B: Effect of child sadness frequency on relative neediness. Outcome variable question wording: How much time, effort, and resources does this child need compared to your other children?
:::

@@

Frequencies of caregiver response to the last instance of child signaling by the cause of the signaling. Categories of causes were based on cluster analysis of more fine-grained cause variables. See @sfig-cause-clustering for cluster dendrogram and @stbl-sumtab-causes for the frequencies of causes and caregiver responses
:::

@@

Effects of selected predictors of the likelihood of positive, neutral, and negative responses to the last time a child became sad or cried when holding continuous variables at their mean and factor variables at their mode. 1 = evidence of the cause in the caregiver report; 0 = no evidence. The endpoints of the lines are dodged slightly for greater visibility. Dot size represents model predictions for each participant given their values of predictor variables. See @sfig-plot-caregiver-punish for the effect of the cause involving punishment
:::

@@

Our bivariate correlations, PCA, MST, and lasso regression analyses all provided a consistent pattern of results: child age was a strong negative predictor of signaling frequency and cost; conflict with the primary caretaker was a positive predictor of signaling frequency and cost; and multiple indices of adversity, such as low neighborhood quality, low household income, and medical problems, were positive predictors of signaling frequency and cost (@fig-PCA,, @fig-mst, @fig-signal-corrplot, @fig-lasso, @fig-signal-effects).

@@

Consistent with the need signaling hypothesis, more frequent sadness was positively associated with perceptions of relative child need within the household (@fig-effect-relativeneed). Higher perceived need, in turn, was associated with higher relative investment (@fig-plot-relative-investment). Crying and tantrums were not associated with higher perceived need, likely due to their high correlation with sadness and the propensity of lasso regression to retain only one of a set of correlated predictors [@zou2005]. Signaling behavior did not predict differences in relative caregiver investment beyond its effect on perceived need [@sfig-plot-invest-coefs], similar to an experimental vignette study that found that the likelihood of costly support was largely mediated by the effect of costly signaling on belief in need [@gaffney2022].

@@

Our results parallel those of previous studies of infants and young children that found that, outside of peaks in infancy and early childhood, the frequencies of crying and temper tantrums decrease with age [@macfarlane1954; @ward1970; @osterman2010; @vermillet2022; @barr1991; @potegal2003; @bhatia1990]. Studies examining crying and tantrums in later childhood and adolensce are rare, especially for crying [@rottenberg2012; @zeifman2013]. Our findings suggest the frequency of crying continues to decrease into childhood, dropping precipitously as children grow older. The frequency of temper tantrums also declined, as did the frequency of child sadness, albeit at a much slower rate (@fig-lasso, @fig-signal-effects). Our results also align with studies examining responses to the crying and fussing of infants that have shown that they often lead to increased support [e.g., breastfeeding, attention, or maintaining proximity, @lin2012; @kushnick2006; @kruger2010; @hewlett1998; @chaudhary2024; @tronick1987], but can also lead  to negative responses like abuse [@zeevenhooven2018; @kurth2011; @howard2006; @soltis2004].

@@

<!-- Move paragrpah below to 461? -->
One possible reason for the negative association of signaling with age, which is also seen in more industrialized settings [@zeifman2013; @potegal2003; @koch2003; @macfarlane1954], is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need [@osterman2010], and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while the energetic cost of crying might be costly for the youngest children, it is likely insignificant for older children. Older children might instead need to weigh the potential social costs of crying or otherwise expressing sadness [@glasberg1981; @glasberg1982; @zeman1996] against the potential increase in social support, which is consistent with our result that the age-related decline in signaling was steepest for crying. The age-related decline in signaling was also generally steeper for males than females (@sfig-signaltype).

@@

Although this study was broader than most in that it examined sadness, crying, and temper tantrums across much of childhood and adolescence, it did not cover all the ways that caregivers might infer need, such as cues of injury. It also did not investigate all potential signals of need. For example, Spanish-speaking mothers reported in focus groups that refusing to do homework was a common tactic. Pained facial and bodily expressions [@vigil2014; @craig2009; @cano2010; @schiavenato2010], self-harm outside of temper tantrums [@hagen2008; @nock2008], and suicidality [@syme2016] have all been suggested to credibly signal need in times of conflicts of interest and may be viable strategies for children, at least at certain ages. Running away has received similar treatment, especially in instances when it comes with substantial risk [@shostak1981] or is seen as a suicide-threat [@fischer1950; @syme2023]. English-speaking mothers in our focus groups emphasized running away and self-harm (e.g., a child slamming their head on the ground). However, this often involved children visiting neighbors or aunts and was much less risky than it would be on mainland Honduras. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean and the presence of dogs, motorcycles, golf carts, and tuk-tuks (auto rickshaws) on the narrow roads typical of much of populated parts of the island. Due to uncertainty on how to conceptualize the risks involved, we did not include running away in our analyses.

The main limitation of this study is that all data come from caregiver reports, and therefore, reflect the biases of caretakers, whatever those might be. For example, caregiver reports of their own responses to recent signals of need might be positively biased. In addition, questions about neighborhood quality lacked a reference, so some caregivers might have compared their neighborhoods to mainland neighborhoods and others to island neighborhoods.
